Open-Source machine learning BANTER acoustic classification of beaked whale echolocation pulses

### 2024-01-12 manuscript/manuscript_files/ReviewResponse_DetailedRebuttal.Rmd

-   [Banter is easy to use open source classifier for bioacoustic data to be used on multiple call types]

-   [Banter can be applied to single call types when multiple detectors are used to identify those individual call types]

-   [how well did banter do at classification?]

I would like to thank the reviewers and editors, especially the Associate Editor, for their review of our manuscript.

-   ***Request to include expanded review of research on the research topic (Introduction).*** The associate editor requested that we expand our Introduction to include work that has been published on the topic. Our intention was to focus on a new application of an existing tool, and I assume that the associate editor would like increased consideration of deep learning methods for acoustic classification. To that end, I have included an additional paragraph in the introduction that includes consideration of these methods and a justification of why despite the incredibly advances in deep learning in recent years, consideration of machine learning methods (such as models based on Random Forest) may be preferred for some classification tasks.

-   **Request to evaluate our results and draw parallels with relevant studies in the peer reviewed literature to provide important context (Discussion).** I included a brief comparison of our results to other machine learning methods to acoustically classify beaked whales, and highlighted how the number of classes and the sample size per class differentiates these studies. ***I also included a brief discussion of how the ease of use of BANTER allows biologists to readily adopt these methods, whereas the barrier to entry for most other machine learning models is steep.*** I hesitate to dive deeper into this discussion as these methods are evolving extremely rapidly, and while deep learning methods are rapidly advancing and are arguably the future of our science, at this point they have requirements that are beyond the scope of many scientists. Our focus is on how to use existing tools to provide reasonable results for imperfect data. I have added an additional statement encouraging the preparation of training sets for the eventual consideration in developing deep learning routines.

-   Provide Context, broader implications of our work, and critical discussion of the uncertainties

-   Improve resolution of Figures

-   Include Research Highlights

### 2024-01-12 manuscript/manuscript.qmd
revisions
revisions to manuscript based on reviewers comments

@@

BANTER (**B**io-**A**coustic eve**NT** classifi**ER**) is a supervised machine learning method originally developed for species identification of dolphin signals from acoustic recordings [@rankin_acoustic_2017]. This hierarchical Random Forest [@breiman_random_2001] event classifier combines information from independent classification models for each call type (e.g., whistles, echolocation pulses, and burst pulses for the dolphin-based model) into a model classifying a full acoustic event, which is defined as a discrete collection of calls during an encounter. The event classifier uses the distribution of species assignment probabilities for each call type, along with any event-level metrics (such as call rate). BANTER is very flexible and can accommodate any number of measures from any number of detectors.

@@

The results for the four study areas varied from a low of 88.7% correct classification in the SAtlantic dataset to a high of 97.5% correct classification in the NAtlantic dataset. All species in all datasets had classification rates well above those expected by chance; however Cuvier's (SAtlantic, Hawaii), Sowerby's (SAtlantic), and Gervais' beaked whales (NAtlantic) had classification rates below 90%.

>>>>

Advances in computing have led to a dramatic increase in the opportunities to use machine learning for bioacoustic classification [@bianco_machine_2019, @mutanu_review_2022, @stowell_computational_2022]. In particular, use of deep neural networks show great promise for large datasets on a wide variety of taxa, including marine mammals [@frainer_automatic_2023, @frasier_automated_2017, @li_model-based_2020, @roman_ruiz_fin_2023, @zhong_beluga_2020]. Neural networks typically require significant sample sizes for all classes, although there are increasingly new tools to enhance data or combined disparate datasets [@dufourq_passive_2022, @nanni_data_2020, @nolasco_learning_2023, @padovese_adapting_2023, @shorten_survey_2019]. There is a significant learning curve to understand how to properly apply and test neural network classification methods on a novel dataset, although development of more user friendly tools are in development [@kahl_birdnet_2021, @kirsebom_meridian_2022]. Despite the great potential for deep learning methods, modification of existing machine learning methods with a lower barrier to entry may be sufficient for many bioacoustic classification tasks.

Random Forest based acoustic classifiers have been found to be relatively easy, efficient, and effective approach to bioacoustic classification [@lopez-baucells_stronger_2019, @ross_random_2014]. BANTER (**B**io-**A**coustic eve**NT** classifi**ER**) is a hierarchical Random Forest [@breiman_random_2001] event classifier originally developed for species identification of dolphin signals from acoustic recordings [@rankin_acoustic_2017]. BANTER combines information from independent classification models for each call type (e.g., whistles, echolocation pulses, and burst pulses for the dolphin-based model) into a model classifying a full acoustic event, which is defined as a discrete collection of calls during an encounter. The event classifier uses the distribution of species assignment probabilities for each call type, along with any event-level metrics (such as call rate). BANTER is very flexible and can accommodate any number of measures from any number of detectors used to identify calls from any species. BANTER is easy to use and is readily available as a stand-alone R package [@archer_banter_2022].

@@

@@

The results for the four study areas varied from a low of 88.7% correct classification in the SAtlantic dataset to a high of 97.5% correct classification in the NAtlantic dataset. All species in all datasets had classification rates well above those expected by chance; however Cuvier's (SAtlantic, Hawaii), Sowerby's (SAtlantic), and Gervais' beaked whales (NAtlantic) had classification rates below 90%. Other studies on machine learning acoustic classification of beaked whales have yielded stronger classification results, although these studies included fewer beaked whale species classes and much larger sample sizes for each species class \[@li_model-based_2020-1, @lebien_species-level_2018, @cohen_identification_2022\]. While there is great room for improvement in our models, and we suspect that modest increases in sample sizes for some species will result in significant improvements to the model results, it is likely that the low sample sizes presented in these datasets would be insufficient to develop and test deep learning models.
